---
title: "Class 9"
author: Lea Sounthong
format: pdf
editor: visual
---

## Class 9: Candy Mini-Project

**2. Importing Candy data**

```{r}
#Importing Candy data
candy_file <- "candy-data.csv"
candy = read.csv (candy_file, row.names = 1)
#Viewing the first 6 rows of the dataset
head(candy)
```

**Q1. How many different candy types are in this dataset?**

```{r}
#Viewing how many different candy types in this dataset
nrow(candy)
```

There are 85 different candy types in this data set

**Q2. How many fruity candy types are in the dataset?**

```{r}
#Viewing sum of fruity candy types
sum(candy$fruity)
```

There are 38 fruity candy types in this dataset.

**2.2 What is your favorite candy?**

```{r}
#Viewing the percentage of people who prefer twix
candy["Twix",]$winpercent
```

```{r}
#Loading dplyr for data functions
library(dplyr)
#Find the win percentage for twix
candy |> 
  filter(row.names(candy)=="Twix") |> 
  select(winpercent)
```

**Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?**

```{r}
candy["Starburst", "winpercent"]
```

The winpercent value of my favorite candy, Starburst, is 67.04.

**Q4. What is the winpercent value for “Kit Kat”?**

```{r}
candy["Kit Kat", "winpercent"]
```

The winpercent value for Kit Kat is 76.77.

**Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?**

```{r}
candy["Tootsie Roll Snack Bars", "winpercent"]
```

The winpercent value for Tootsie Roll Snack Bars is 49.65.

```{r}
library("skimr")
skim(candy)
```

**Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?**

The winpercent column seems to be on a different scale to the majority of the other columns in the dataset. Most of columns in the data set range from 0 to 1 but the winpercent is ranging above 10.

**Q7. What do you think a zero and one represent for the candy\$chocolate column?**

A zero in the candy\$chocolate column would represent that the candy does not have chocolate, but a 1 in this column would indicate that the candy does have a chocolate.

**3. Exploratory Analysis**

**Q8. Plot a histogram of winpercent values using both base R an ggplot2.**

```{r}
#Plotting a histogram of winpercent values using base R
hist(candy$winpercent,
     main = "Distribution of Winpercent",
     xlab = "Winpercent",
     col = "purple",
     border = "black")
```

```{r}
#Loading in ggplot2
library(ggplot2)
#Plotting a histogram of winpercent values using ggplot2
ggplot(candy, aes(x = winpercent)) +
  geom_histogram(binwidth = 2, fill = "purple", color = "black") +
  labs(title = "Distribution of Winpercent",
       x = "Winpercent",
       y = "Count")
```

**Q9. Is the distribution of winpercent values symmetrical?**

The distribution of winpercent values is not symmetrical since it looks slightly right skewed, and there is a longer tail toward the higher winpercent values.

**Q10. Is the center of the distribution above or below 50%?**

The center of the distribution is below 50%

**Q11. On average is chocolate candy higher or lower ranked than fruit candy?**

```{r}
#Generating Welch two sample t-test to compare the mean winpercent 
#for chocolate and fruity candies 
t.test(candy$winpercent[candy$chocolate == 1],
       candy$winpercent[candy$fruity == 1])
```

On average, the chocolate candy has a higher winpercent of 60.92, compared to fruity candy of 44.12.

**Q12. Is this difference statistically significant?** The p-value of the t=test was 2.871e-08, which is much less than 0.05, therefore the difference is statistically significant.

**4. Overall Candy Rankings**

**Q13. What are the five least liked candy types in this set?**

```{r}
#Sorting dataset by winpercent from lowest to highest and displaying first 5 rows
head(candy[order(candy$winpercent),], 5)
```

```{r}
#Using dplyr to sort dataset by winpercent from lowest to highest
#and display first 5 rows 
candy |> 
  arrange(winpercent) |>
  head(5)
```

The five least liked candy types in this set are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.

**Q14. What are the top 5 all time favorite candy types out of this set?**

```{r}
#Sorting the dataset by winpercent from highest to lowest and displaying first 5 rows 
head(candy[order(-candy$winpercent),], 5)
```

```{r}
#Using dplyr to sort dataset by winpercent from highest to lowest 
#and display first 5 rows 
candy |> 
  arrange(desc(winpercent)) |>
  head(5)
```

The top 5 all time favorite candy types out of this set are Reese's Peanut Butter cup, Reese's Miniatures, Twix, Kit Kat, and Snickers.

Overall, I prefer to use the dplyr approach since it seems more organized and easier to read.

**Q15. Make a first barplot of candy ranking based on winpercent values.**

```{r}
#Loading ggplot2
library(ggplot2)
#Generating basic barplot of candy ranked based on the winpercent
ggplot(candy) +
  aes(x = winpercent, 
      y = rownames (candy)) +
  geom_col()
```

**Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?**

```{r}
#Loading ggplot2
library(ggplot2)
#Generating barplot with candies sorted by the winpercent
ggplot(candy) +
  aes(x = winpercent,
      y = reorder(rownames(candy), winpercent)) +
  geom_col()
```

**4.0.1 Time to add some useful color**

```{r}
#Creating vector of colors 
#Default all black, one for each candy
my_cols=rep("black", nrow(candy))
#Changing color for candies with chocolate to the color chocolate
my_cols[as.logical(candy$chocolate)] = "chocolate"
#Changing color for candies that are bars to brown
my_cols[as.logical(candy$bar)] = "brown"
#Changing color for candies that are fruity to pink 
my_cols[as.logical(candy$fruity)] = "pink"
#Generating barplot with colors based on the type of candy
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

**Q17. What is the worst ranked chocolate candy?**

The worst ranked chocolate candy is Sixlets.

**Q18. What is the best ranked fruity candy?**

The best ranked fruity candy is Starburst.

**5. Taking a look at pricepercent**

```{r}
#Loading ggrepel
library(ggrepel)
#Generating plot of winpercent vs pricepercent
ggplot(candy) +
  aes( x = winpercent, 
       y = pricepercent, 
      label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
  
```

**Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?**

Reese's Miniatures is the highest ranked in terms of winpercent for the least money, so it offers the most bang for your buck.

**Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?**

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

The five most expensive candies are Nik L Nip, Nestle Smarties, Ring Pop, Hershey's Krackel, and Hershey's Milk Chocolate. The least popular is the Nik L Nip, since it has the lowest winpercent.

**5.1 Optional**

**Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().**

```{r}
#Generating barplot and order candies by
ggplot(candy)+
  aes(x = pricepercent,
      y = reorder(rownames(candy), pricepercent)) +
  geom_col() +
  labs(y = "Candy", x = "Pricepercent")
```

```{r}
#Generating a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

**6. Exploring the correlation structure**

```{r}
#Loading corrplot package
library(corrplot)
```

```{r}
#Generating a correlation matrix of the candy data 
cij <- cor(candy)
corrplot(cij)
```

**Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?**

The two variables that are anti-correlated are chocolate and fruity.

**Q23. Similarly, what two variables are most positively correlated?**

The two variables that are most positively correlated are chocolate and bar.

**7. Principal Component Analysis**

```{r}
#Applying PCA on the candy dataset with scaling
pca <- prcomp(candy, scale = TRUE)
#Viewing summary of PCA results
summary(pca)
```

```{r}
#Plotting PC1 vs PC2
plot(pca$x[,1:2])
```

```{r}
#Changing the plotting character and adding color
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
#Making a new data fram with PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
#Generating PCA plot using ggplot
p <- ggplot(my_data) +
  aes (x = PC1,
       y = PC2,
       size = winpercent/100,
       text = rownames(my_data)) +
  geom_point(col = my_cols)

#Viewing the plot p
p
```

```{r}
#Labeling plot with non overlapping candy names using ggrepel
library(ggrepel)

p + 
  geom_text_repel(aes(label = rownames(my_data)),
                  size=3.3, 
                  col=my_cols, 
                  max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r, fig.align="center", out.height = "4.5in"}
#Using max.overlaps to allow more overlapping lables 
library(plotly)

library(webshot2)
pp <- ggplotly(p)
htmlwidgets::saveWidget(pp, "plot.html", selfcontained = TRUE)
webshot2::webshot("plot.html", file = "plot.png")
```

```{r}
#Plotting the loadings for PC1
ggplot(pca$rotation) +
  aes(x = PC1,
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col() +
  labs (x = "PC1", y = "Candy")
```

**Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?**

Original variables that are picked up strongly by PC1 in this positive direction are fruity, pluribus, and hard. This makes sense because these variables describe the features of fruity and non chocolate candies, which are different than chocolate related candies. This relationship was highlighted previously in the correlation plot, since fruity and chocolate were negatively correlated.

**Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?**

A "winning" candy would be a chocolate based candy with added textures such as caramel or peanuts. Visualizations, including bar and scatter plots, highlighted their higher popularity since they have higher winpercent values than fruity candies. Correlation analysis demonstrated how chocolate and fruity candies are strongly anti-correlated, suggesting that these are distinct categories with different popularity levels. PCA further supported this by grouping the most popular candies within the chocolate region. 