---
title: "Class 08"
author: Lea Sounthong
format: pdf
editor: visual
---

## Class 8 Mini-Project

**Preparing the data**

```{r}
# Save your input data file into your Project directory
fna.data <- "WisconsinCancer.csv"

# Inputting the data and store as wisc.df
wisc.df <- read.csv(fna.data, row.names=1)
```

```{r}
#Examining input data to ensure column names are set correctly
head(wisc.df, 4)
```

```{r}
#Using -1 to remove the first column
wisc.data <- wisc.df[,-1]
```

```{r}
# Create diagnosis vector for later 
diagnosis <- as.factor(wisc.df$diagnosis) 
```

**Exploratory data analysis**

**Q1. How many observations are in this dataset?**

```{r}
#Using nrow to show the number of rows int he dataset
nrow(wisc.df)
```

There are 569 observations in this data set

**Q2. How many of the observations have a malignant diagnosis?**

```{r}
#Using table() to count how many times each category appears in diagnosis vector
table(diagnosis)
```

There are 212 malignant diagnoses.

**Q3. How many variables/features in the data are suffixed with \_mean?**

```{r}
#Using grep() to find the column names that end with _mean. 
#Using legnth() to find the amount of matches
grep("_mean$", colnames(wisc.df), value = TRUE)
length(grep("_mean$", colnames(wisc.df)))
```

10 of the variables/features in the data are suffixed with \_mean.

**3. Principal Component Analysis**

**Performing PCA**

```{r}
#Using -1 to remove the first column
wisc.data <- wisc.df[,-1]
# Checking column means and standard deviations
colMeans(wisc.data)
apply(wisc.data,2,sd)
```

```{r}
#Perfoming PCA on wisc.data
wisc.pr <- prcomp(wisc.data, scale = TRUE)
```

```{r}
#Looking at the PCA summary results
summary(wisc.pr)
```

**Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?**

The proportion of the original variance captured by the PC1 is 0.4427, or 44.27%.

**Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?**

Looking at the cumulative proportion, PC3 is the first time the total variance explained exceeds 70%, since it reaches 0.72636 or \~72%, therefore only 3 principal components are required.

**Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?**

Looking at the cumulative proportion, PC7 is the first time the total variance explained exceeds 90%, since it reaches 0.91010 or \~91%, therefore only 7 principle components are required.

**Interpreting PCA results**

```{r}
biplot(wisc.pr)
```

**Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?**

What stands out to me about this biplot is there there are many overlapping numbers and labels. This makes it hard to understand.

```{r}
#Loading in ggplot2
library(ggplot2)
#Generating a scatter plot of the first two principal components
ggplot(wisc.pr$x) +
  #Setting PC1 for x-axis and PC2 for y-axis and coloring points by diagnosis
  aes(PC1, PC2, col = diagnosis) +
  #Adding points for each observation
  geom_point()
  
```

**Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?**

```{r}
#Generating scatter plot of PC1 and PC3
ggplot(wisc.pr$x) +
#Setting P1 to x-axis and PC3 to y-axis and color points by diagnosis
  aes(PC1, PC3, col = diagnosis) +
  #Adding points for each observation
  geom_point()
```

From the plots I notice that PC1 is better at separating malignant from benign samples in both plots. However, the PC1 vs PC3 plot has more overlap in points than PC1 vs PC2, suggesting that PC3 is less effective at separating between the malignant and benign diagnoses.

**Variance Explained**

```{r}
#Calculating the variance of each principal component 
#by squaring the sdev component of wisc.pr
pr.var <- wisc.pr$sdev^2

#Viewing the first few values
head(pr.var)
```

```{r}
#Calculating the variance explained by each principal component 
#Dividing each PC variance by the total variance to get proportion explained
pve <- pr.var / sum(pr.var)
#Plotting the variance explained for each principal component
plot(c(1,pve), 
     #Labeling x-axis and y-axis
     xlab = "Principal Component",
     ylab = "Proportion of Variance Explained",
     #Setting the y-axis limits from 0 to 1
     ylim = c(0,1),
     #Plotting both the points with a trend line
     type = "o")
```

```{r}
# Alternative scree plot of the same data, note data driven y-axis
#Generating a barplot 
barplot(pve, ylab = "Percent of Variance Explained",
    #Labeling each bar as PC1, PC2, etc. using names.arg
     names.arg=paste0("PC",1:length(pve)), 
    #Rotating x-axis labels so they are vertical and removing default y-axis
    las=2, axes = FALSE)
#Adding custom y-axis on left side
axis(2, 
     # Puts marks on each PVE value
     at=pve, 
     #Converting proportions to percentages and rounding them
     labels=round(pve,2)*100 )
```

```{r}
## ggplot based graph
#install.packages("factoextra")
library(factoextra)
#Generating scree plot
fviz_eig(wisc.pr, addlabels = TRUE)
```

**Communicating PCA results**

**Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr\$rotation\[,1\]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?**

```{r}
#Retrieving the PC1 loading for the feature concave.points_mean 
#to see its contribution to PC1
wisc.pr$rotation["concave.points_mean", 1]
```

```{r}
#Veiwing all PC1 loadings and sort by size
sort(wisc.pr$rotation[,1], decreasing = TRUE)
```

On the PC1, the loading value for concave.points_mean is -0.2608538. This suggests that this feature contributes strongly to PC1. There are no features with larger contributions than this one.

**4.Hierarchial Clustering**

```{r}
#Scaling the wisc.data
data.scaled <- scale(wisc.data)
```

```{r}
#Calculating Euclidean distances between observations
data.dist <- dist(data.scaled)
```

```{r}
#Creating a hierarchial clustering model using complete linkage
wisc.hclust <- hclust(data.dist, method = "complete")
```

**Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?**

```{r}
#Plotting the hierarchial clustering dendrogram
plot(wisc.hclust)
abline(h=19, col = "red", lty = 2)
```

The height at which the clustering model has 4 clusters is around 19.

**Selecting number of clusters**

```{r}
#Cuting the hierarchial clustering tree to 4 clusters
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
#zGenerating table comparing cluster assignments to true labels
table(wisc.hclust.clusters, diagnosis)
```

**Q11. OPTIONAL: Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 6? How do you judge the quality of your result in each case?**

```{r}
#Trying k = 2 to 6 to comapre each diagnosis
for (k in 2:6) {
  clusters <- cutree(wisc.hclust, k = k)
  print(table(clusters, diagnosis))
}
```

The best match occurred at k = 5, where the clusters are more pure with less mixing than the other clusters. The quality of clustering was measured by comparing the clusters to diagnosis labels. The other clustering such as k = 2 and k = 3 did not have as good separation because these clusters are highly mixed, while k = 6 is split into many clusters without improving the separation between the diagnoses.

**Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.**

The method that gives me my favorite results for the same data.dist dataset is ward.D2 because it makes the clusters more separated and organized without mixing the diagnoses of benign and malignant samples. This makes the clusters easier to interpret, and makes more logical sense for this biological scenario compared to the single, complete, or average methods.

**5. Combining Methods**

**Comparing on PCA Results**

```{r}
#Using PC scores from wisc.pr$x to keep the minimum PCs needed for 90% variability 
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method = "ward.D2")
#Plotting the hierarchial clustering dendrogram for wisc.pr.clust
plot(wisc.pr.hclust)
abline(h=19, col = "red", lty = 2)
```

```{r}
#Cutting the dendrogram to 2 clusters
grps <- cutree (wisc.pr.hclust, k = 2)
#Counting how many observatings are in each cluster
table(grps)
#Comparing the clusters to diagnoses
table(grps, diagnosis)
```

```{r}
#Plotting PC1 vs PC2 and color each by its cluster assignment 
ggplot(wisc.pr$x) +
  aes(PC1, PC2) +
  geom_point (col = grps)
```

**(Skipping the 3D part for pdf report)**

```{r}
#Using the distance along the first 7 principal components
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method = "ward.D2")
#Cutting PCA-based clustering into 2 clusters
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k = 2)
```

**Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?**

```{r}
#Comparing PCA-based clustering to actual diagnosis
table(wisc.pr.hclust.clusters, diagnosis)
```

This hclust model with two clusters separate out the two diagnoses well since one cluster is mostly malignant (188 M vs 28 B) while the other is mostly benighn (329 B Vs 24 M). This demonstrates strong alignment with the actual diagnosis.

**Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.**

```{r}
#Comparing clustering without PCA
table(wisc.hclust.clusters, diagnosis)
#Comparing clustering with PCA
table(wisc.pr.hclust.clusters, diagnosis)
```

The PCA based clustering model separates the malignant and benign samples more effectively than the non PCA based model. The PCA based clustering model shows clearer clusters with less overlap between the two diagnoses.

**6. Sensitivity/Specificity**

**Q15. OPTIONAL: Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?**

The PCA based clustering model had the best specificity with 92% and the best sensitivity with around 89%. This demonstrates that the model can effectively identify benign and malignant samples.

**7. Prediction**

```{r}
#Inputting url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
#Plotting PC1 vs PC2 from the original data and coloring points by cluster assignment
plot(wisc.pr$x[,1:2], col=grps)
#Overlaying new patient samples in blue and label them 1 and 2
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

**Q16. Which of these new patients should we prioritize for follow up based on your results?**

Based on these results, the new patient we should prioritize is patient 1 since they are in the cluster associated with malignant samples. Patient 2 is in the benign cluster, so this case has less urgency.
