---
title: "Class 12"
author: Lea Sounthong
format: pdf
editor: visual
---

## Transcriptomics and the analysis of RNA-Seq data

**3. Import countData and colData**


```{r}
#Read counts file and metadata file
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")
```

```{r}
#View first rows of counts
head(counts)
```

```{r}
#View first rows of metadata
head(metadata)
```

**Q1. How many genes are in this dataset?**

```{r}
#Finding number of genes (rows)
nrow(counts)
```

There are 38694 genes.

**Q2. How many ‘control’ cell lines do we have? **

```{r}
#Finding number of 'control' cell lines
View(metadata)
```

There are 4 control cell lines.

**4. Toy differential gene expression**

```{r}
#Obtaining the control samples
control <- metadata[metadata[,"dex"]=="control",]
#Obtaining counts for control samples
control.counts <- counts[ ,control$id]
#Calculating the mean count per gene
control.mean <- rowSums( control.counts )/4 
#Viewing the first values
head(control.mean)
```

```{r}
library(dplyr)
#Obtaining treated samples
control <- metadata %>% filter(dex=="control")
#Obtaining counts for treated samples
control.counts <- counts %>% select(control$id) 
#Calculating the mean count per gene
control.mean <- rowSums(control.counts)/4
#Viewing the first values 
head(control.mean)
```

**Q3. How would you make the above code in either approach more robust? Is there a function that could help here?**

The function that could help here is rowMeans() which could make it more robust because it calculates the mean across the selected samples automatically. So if more saples were added, the values obtained would still be correct.

**Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)**

```{r}
#Obtaining treated samples
treated <- metadata[metadata[,"dex"] =="treated",]
#Obtaining counts for treated samples 
treated.counts <- counts[, treated$id]
#Calculating the mean count per gene
treated.mean <- rowSums( treated.counts) / 4 
#Viewing first values
head(treated.mean)
```

```{r}
#Combining metacount data 
meancounts <- data.frame(control.mean, treated.mean)
```

```{r}
#Sum mean counts for each group
colSums(meancounts)
```

**Q5 (a). Create a scatter plot showing the mean of the treated samples against the mean of the control samples. Your plot should look something like the following.**

```{r}
#Generating scatter plot showing mean of treated samples vs mean of control samples
plot(meancounts[,1], meancounts[,2],
     xlab = "Control",
     ylab = "Treated")
```

**Q5 (b).You could also use the ggplot2 package to make this figure producing the plot below. What geom_?() function would you use for this plot?**

```{r}
#Loading ggplot2 package
library(ggplot2)
#Generating scatterplot with ggplot2
ggplot(meancounts, aes(x = control.mean, y = treated.mean)) +
  geom_point()
```

The geom_() function would be geom_point().

**Q6. Try plotting both axes on a log scale. What is the argument to plot() that allows you to do this?**

```{r}
#Generating scatter plot with both axes on a log scale
#Adding +1 to prevent zero values 
plot(meancounts[, 1] + 1, meancounts[, 2] +1,
     xlab ="log Control",
     ylab = "log Treated",
     log = "xy")
```

The argument to plot() that allows me to do this is log = "xy"

```{r}
#Calculating log2foldchange and adding it to meancounts data.frame
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
#Viewing results
head(meancounts)
```

```{r}
#Filtering data to remove genes that have zero expression
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
#Viewing results
head(mycounts)
```

**Q7. What is the purpose of the arr.ind argument in the which() function call above? Why would we then take the first column of the output and need to call the unique() function?**

The purpose of the arr.ind=TRUE argument is to tell which() to return the positions of TRUE values (so values that are 0) as rows and columns. We then select the first column to identify genes with zero counts and use unique() so each gene only shows up one time.

```{r}
#Filtering the dataset both ways to see how any genes are up or down regulated
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

**Q8. Using the up.ind vector above can you determine how many up regulated genes we have at the greater than 2 fc level? **

```{r}
#Finding the number of up regulated genes
paste("Up:", sum(up.ind))
```

There are 250 up regulated genes

**Q9. Using the down.ind vector above can you determine how many down regulated genes we have at the greater than 2 fc level?**

```{r}
#Finding the number of down regulated genes
paste("Down:", sum(down.ind))
```

There are 367 down regulated genes

**Q10. Do you trust these results? Why or why not?**

I do not trust these results because they are only based on fold change and they do have not performed statistical testing yet, so the this alone is not able to deterine if the gene expression differences are statistically significant.

**5. Setting up for DESeq**

```{r}
library(DESeq2)
citation("DESeq2")
```

```{r}
#Using the DESeqDataSetFromMatric() function to build required DESeqDataSet object 
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```

**6. Principal Component Analysis (PCA)**

```{r}
#Transforming the raw count data where the variance is approximately 
#independent of the mean
vsd <- vst(dds, blind = FALSE)
#Generating PCA plot grouped by dex treatment
plotPCA(vsd, intgroup = c("dex"))
```

```{r}
#Obtaining the PCA plot data
pcaData <- plotPCA(vsd, intgroup=c("dex"), returnData=TRUE)
#Viewing the first rows of PCA data
head(pcaData)

```

```{r}
#Calculating percent variance per PC for the plot axis labels
percentVar <- round(100 * attr(pcaData, "percentVar"))
```

```{r}
#Generating PCA plot using ggplot2
ggplot(pcaData) +
#Setting PCA Axes and color by the treatment group
  aes(x = PC1, y = PC2, color = dex) +
#Adding saple points and labeling the x and y axis with percent value
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
#Keeping the axes proportional
  coord_fixed() +
#Black and white theme
  theme_bw()
```

**7. DESeq analysis**

#Viewing results before running DESeq
results(dds) gave an error 


```{r}
#Running DESeq2 differential expression analysis 
dds <- DESeq(dds)
```

```{r}
#Extracting DESeq2 results and veiewing the results
res <- results(dds)
res
```

```{r}
#Converting results object to a dataframe and viewing in a data viewer
res.df <- as.data.frame(res)
View(res.df)
```

```{r}
#Viewing summary of DESeq2 results
summary(res)
```

```{r}
#Obtaining DESeq2 results with adjusted p-value cutoff of 0.05
res05 <- results(dds, alpha=0.05)
#Viewing summary of results
summary(res05)
```

**8. Adding annotation data**

```{r}
#Loading packages
library("AnnotationDbi")
library("org.Hs.eg.db")
```

```{r}
#Viewing annotation columns that are available 
columns(org.Hs.eg.db)
```

```{r}
#Adding gene symbols to results table
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res), #Our gene IDs
                     keytype="ENSEMBL",   #The current format of our gene IDs
                     column="SYMBOL",     #Convert to gene symbols
                     multiVals="first")   #Use first match if multiple exist
#Viewing first rows of results 
head(res)
```

**Q11. Run the mapIds() function two more times to add the Entrez ID and UniProt accession and GENENAME as new columns called res$entrez, res$uniprot and res$genename.** 

```{r}
#Adding Entrez gene IDs
res$entrez <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="ENTREZID",
                     keytype="ENSEMBL",
                     multiVals="first")
#Adding UniProt accession IDs
res$uniprot <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="UNIPROT",
                     keytype="ENSEMBL",
                     multiVals="first")
#Adding GENENAME gene IDs
res$genename <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="GENENAME",
                     keytype="ENSEMBL",
                     multiVals="first")
#Viewing the first rows of updated table
head(res)
```

```{r}
#Ordering results by adjusted p-value
ord <- order( res$padj )
#View(res[ord,])
head(res[ord,])
```

```{r}
#Exporting ordered results table
write.csv(res[ord,], "deseq_results.csv")
```

**9. Data Visualization**

```{r}
#Generating volcano plot of fold change vs p-value
plot( res$log2FoldChange,  -log(res$padj), 
      xlab="Log2(FoldChange)",
      ylab="-Log(P-value)")
```

```{r}
#Generating volcano plot and adding guidelines with abline() and color 
#Highlighting genes that have padj<0.05 and the absolute log2FoldChange>2 

plot( res$log2FoldChange,  -log(res$padj), 
 ylab="-Log(P-value)", xlab="Log2(FoldChange)")

# Add some cut-off lines
abline(v=c(-2,2), col="darkgray", lty=2)
abline(h=-log(0.05), col="darkgray", lty=2)
```

```{r}
#Creating default color vector for volcano plot 
mycols <- rep("gray", nrow(res))
#Coloring genes with large fold-change red
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red" 

#Identifying highly significant genes
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
#Coloring significant genes blue
mycols[ inds ] <- "blue"

#Generating volcano plot with custom colors
plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

#Adding fold-change cutoff lines
abline(v=c(-2,2), col="gray", lty=2)
#Adding p-value cutoff line
abline(h=-log(0.1), col="gray", lty=2)
```

```{r}
#Loading Enhanced Volcano library
library(EnhancedVolcano)
#Converting results object to a data frame
x <- as.data.frame(res)

#Generating labeled volcano plot 
EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')
```

