---
title: "Class05: Data Visualization with GGPLOT"
author: Lea Sounthong
format: pdf
editor: visual
---

**Class Work:**

# Title Here

## Subtitle

###Subtitle

**Stuff in bold** not in bold *something italics*

> Indent text

```{r}
#Generating numbers from 1-5 and adding 1 to each number
x<-1:5
x+1
```

Plot from class:

```{r}
#Generating scatter plot using the mpg dataset 
library(ggplot2)
ggplot(data=mpg) +
  #Setting variables for axes and grouping 
  aes(x=displ, y=hwy, shape=factor(cyl)) +
  #Adding points to the plot
  geom_point() + 
  #Split plot into panels by cylinder number 
  facet_wrap(~cyl) +
  #Generating simple theme
  theme_bw() +
  #Generating labels and titles 
  labs(title="Cars", subtitle= "hwy mileage", 
       caption="my data", x="displacement", y="Mileage")
```

**Q1. For which phases is data visualization important for our scientific workflows?**

All of the above

**Q2. TRUE or FALSE? The ggplot2 package comes already installed with R?**

FALSE



**Part 5**

**Q. Which plot types are typically NOT used to compare distributions of numeric variables?**

Network graphs

**Q. Which statement about data visualization with ggplot2 is incorrect?**

ggplot2 is the only way to create plots in R

**Part 6. Creating Scatter Plots**

**Q. Which geometric layer should be used to create scatter plots in ggplot2?**

geom_point()

**Generating scatter plot for part 6**

```{r}
#Loading ggplot2 for creating plots
library(ggplot2)
#Create a scatter plot of the cars dataset
ggplot(cars)+
  #Define which variables used for x and y axes and making points blue 
  aes(x=speed,y=dist)+geom_point(col="Blue")
```

**Q. In your own RStudio can you add a trend line layer to help show the relationship between the plot variables with the geom_smooth() function?**

```{r}
#Generating a trend line layer to show relationship between the plot variables 
library(ggplot2)
ggplot(cars) + 
  aes(x = speed, y = dist) +
  geom_point() +
  geom_smooth()
```

**Q. Argue with geom_smooth() to add a straight line from a linear model without the shaded standard error region?**

```{r}
#Generating scatter plot of cars data with linear trend line
library(ggplot2)
ggplot(cars) + 
  aes(x = speed, y = dist) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

**Q. Can you finish this plot by adding various label annotations with the labs() function and changing the plot look to a more conservative “black & white” theme by adding the theme_bw() function:**

```{r}
#Adding various label annotations and changing the plot to black and white theme
library(ggplot2)
ggplot(cars) + 
  aes(x = speed, y = dist) +
  geom_point() +
  labs(title="Speed and Stopping Distances of Cars",
       x="Speed (MPH)",
       y="Stopping Distance (ft)",
       subtitle = "Your informative subtitle text here",
       caption="Dataset: 'cars'") +
  geom_smooth(method="lm", se=FALSE) +
  theme_bw()
```

**Adding more plot aesthetics through aes()**

```{r}
#Loading in gene expression dataset from online file 
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt" 
#Read the data into R and preview it
genes <- read.delim(url)
head(genes)
```

**Q. Use the nrow() function to find out how many genes are in this dataset. What is your answer?**

```{r}
#Counting total number of genes in the dataset
nrow(genes)
```

**Q. Use the colnames() function and the ncol() function on the genes data frame to find out what the column names are (we will need these later) and how many columns there are. How many columns did you find?**

```{r}
#Displaying column names in the dataset
colnames(genes)
#Counting how many columns are present 
ncol(genes)
```

**Q. Use the table() function on the State column of this data.frame to find out how many ‘up’ regulated genes there are. What is your answer?**

```{r}
#Counting how many genes are in each category
table(genes$State)
```

**Q. Using your values above and 2 significant figures. What fraction of total genes is up-regulated in this dataset?**

```{r}
#Calculating the percentage of each gene category
round( table(genes$State)/nrow(genes) * 100, 2)
```

**Q. Complete the code below to produce the following plot**

```{r}
#Generating scatter plot of genes data set comparing Condition 1 and Condition 2
ggplot(genes) + 
  aes(x=Condition1, y=Condition2) +
  geom_point()
```

```{r}
#Adding State column to see the difference in expression values between conditions 
ggplot(genes) + 
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point()
```

```{r}
#Adding another layer to explicitly specify color scale
ggplot(genes) + 
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point() +
  scale_colour_manual( values=c("blue", "gray", "red"))
```

**Q. Nice, now add some plot annotations to the p object with the labs() function**

```{r}
#Adding plot annotations using labs()
ggplot(genes) + 
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point() +
  scale_colour_manual( values=c("blue", "gray", "red")) +
  labs(title="Gene Expression Changes Upon Drug Treatment",
       x="Control (no drug)",
       y="Drug Treatment")
```

**Part 7. Going Further**

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
# install.packages("dplyr") 
library(dplyr)

gapminder_2007 <- gapminder %>% filter(year==2007)
```

**Q. Complete the code below to produce a first basic scater plot of this gapminder_2007 dataset:**

```{r}
#Generating scatter plot of gapminder_2007 dataset 
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp) +
  geom_point()
```

```{r}
#Making the points slightly more transparent with alpha=0.4
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp) +
  geom_point(alpha=0.4)
```

```{r}
#Mapping the continent variable to the point color aesthetic and the population pop
#through the point size argument to aes() to obtain plot with 4 different variables
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) +
  geom_point(alpha=0.5)
```

```{r}
#Generating plot and color the point by the numeric variable population pop
ggplot(gapminder_2007) + 
  aes(x = gdpPercap, y = lifeExp, color = pop) +
  geom_point(alpha=0.8)
```

```{r}
#Plotting GDP per capita vs. the life expectancy and set the point size
#based on the population of each country
ggplot(gapminder_2007) + 
  aes(x = gdpPercap, y = lifeExp, size = pop) +
  geom_point(alpha=0.5)
```

```{r}
#Reflect the actual population differences by the point size and 
#using the scale_size_area() function 
ggplot(gapminder_2007) + 
  geom_point(aes(x = gdpPercap, y = lifeExp,
                 size = pop), alpha=0.5) + 
  scale_size_area(max_size = 10)
```

**Q. Can you adapt the code you have learned thus far to reproduce our gapminder scatter plot for the year 1957? What do you notice about this plot is it easy to compare with the one for 2007?**

```{r}
#Generating gapminder scatter plot for year 1957
gapminder_1957 <- gapminder %>% filter(year==1957)

ggplot(gapminder_1957) + 
  aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop) +
  geom_point(alpha=0.7) + 
  scale_size_area(max_size = 10) 
```

**Q. Do the same steps above but include 1957 and 2007 in your input dataset for ggplot(). You should now include the layer facet_wrap(~year) to produce the following plot:**

```{r}
#Generating scatter plot to include both 1957 and 2007

gapminder_1957 <- gapminder %>% filter(year==1957 | year==2007)

ggplot(gapminder_1957) + 
  geom_point(aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop), alpha=0.7) + 
  scale_size_area(max_size = 10) +
  facet_wrap(~year)
```

**Part 8: Bar Charts**

```{r}
#Generating the number of people in the five biggest countries by populatiion in 2007
gapminder_top5 <- gapminder %>% 
  filter(year==2007) %>% 
  arrange(desc(pop)) %>% 
  top_n(5, pop)

gapminder_top5

```

```{r}
#Generating bar chart with gapminder_top5 dataset
ggplot(gapminder_top5) + 
  geom_col(aes(x = country, y = pop))
```

**Q Create a bar chart showing the life expectancy of the five biggest countries by population in 2007.**

```{r}
# Generating bar chart with gapminder_top5 dataset
ggplot(gapminder_top5) +
  geom_col(aes(x = country, y = lifeExp))
```

```{r}
#Filling the bars with color
ggplot(gapminder_top5) + 
  geom_col(aes(x = country, y = pop, fill = continent))
```

```{r}
#Using numeric variable (life expectancy) to fill color
ggplot(gapminder_top5) + 
  geom_col(aes(x = country, y = pop, fill = lifeExp))
```


**Q. Plot population size by country. Create a bar chart showing the population (in millions) of the five biggest countries by population in 2007.**

```{r}
#Generating bar chart showing population (in millons) of the five biggest countries by population in 2007
ggplot(gapminder_top5) +
  aes(x=country, y=pop, fill=gdpPercap) +
  geom_col()
```

```{r}
#Changing the order of the bars
ggplot(gapminder_top5) +
  aes(x=reorder(country, -pop), y=pop, fill=gdpPercap) +
  geom_col()
```

```{r}
#Filling just by country
ggplot(gapminder_top5) +
  aes(x=reorder(country, -pop), y=pop, fill=country) +
  geom_col(col="gray30") +
  guides(fill="none")
```

**Flipping bar charts**

```{r}
#Loading dataset USArrests
head(USArrests)
```

```{r}
#Generating bar chart of USArrests
USArrests$State <- rownames(USArrests)
ggplot(USArrests) +
  aes(x=reorder(State,Murder), y=Murder) +
  geom_col() +
  coord_flip()
```

```{r}
#Plotting USArrests by using points and line segments 
ggplot(USArrests) +
  #Mapping variables
  aes(x=reorder(State,Murder), y=Murder) +
  geom_point() +
  #Adding verticle line segments from 0 up to each state's murder rate
  geom_segment(aes(x=State, 
                   xend=State, 
                   y=0, 
                   yend=Murder), color="blue") +
  #Flip the axes so states are on the y=axis
  coord_flip()
```

**Part 9. Extensions: Animation**

```{r}
library(gapminder)
library(gganimate)

# Setup nice regular ggplot of the gapminder data
ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  # Facet by continent
  facet_wrap(~continent) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  shadow_wake(wake_length = 0.1, alpha = FALSE)
```

**Part 10: Combining Plots**

```{r}
library(patchwork)

# Setup some example plots 
p1 <- ggplot(mtcars) + geom_point(aes(mpg, disp))
p2 <- ggplot(mtcars) + geom_boxplot(aes(gear, disp, group = gear))
p3 <- ggplot(mtcars) + geom_smooth(aes(disp, qsec))
p4 <- ggplot(mtcars) + geom_bar(aes(carb))

# Use patchwork to combine them here:
(p1 | p2 | p3) /
      p4
```


